Update of the ozone dry deposition in the OsloCTM3 v1.0

### 2018-10-05 DryDep.tex
Finished first draft of introduction.

\begin{abstract}
  Since the industrial revolution, tropospheric background ozone concentrations have been increasing strongly mainly in the northern hemisphere. In recent years, the number of episodes of peak concentrations especially in summertime has been decreasing in North America and Europe due to the implementation of air quality regulations. At the same time, fast developing countries, like e.g., China or India, saw a significant increase in ozone related air pollution. High concentrations of ozone in ambient air are hazardous not only to humans but to the ecosystem in general. The impact of ozone damage on vegetation and agricultural plants in particular may even affect food security in the future. Most chemistry-transport models (CTMs) and chemistry-climate models (CCMs) tend to overestimate surface ozone values. This may be partly due to uncertainties in the distribution and amount of precursor species and stratosphere-troposphere exchange as well as scavenging processes removing ozone from the atmosphere. Removal of any gas through gravitational settling or by uptake by plants and soil is referred to as dry deposition. The process of dry deposition is important for predicting surface ozone concentrations and understanding the observed amount and increase of tropospheric background ozone. Dry deposition velocities are typically calculated following a resistance-analogous approach wherein aerodynamic, quasi laminar, and canopy resistances are key components, but these conceptual dry deposition velocities are hard to measure explicitly. Following the European Monitoring and Evaluation Programme (EMEP) MSC-W chemistry transport model, we compute canopy resistance from stomatal conductance. In this paper, we present an update on the dry deposition scheme implemented in the OsloCTM3v1.0 and evaluate the model with respect to ozone and different parameters in the stomatal conductance parameterization, e.g., photosynthetic photon flux density (PPFD) and phenology. Furthermore, we investigate the impact on modeled ozone concentrations, both at the surface and in the free troposphere. We show that ozone dry deposition in the new version (v.1.1) decreases with respect to the previous version (v1.0), leading to an increase in surface ozone of up to 40\,\unit{\%}. We also put our results in context to recent multi-model studies as well as satellite data.
\end{abstract}

@@

Tropospheric ozone is mainly produced in situ in complex photochemical cycles involving precursor gases such as carbon monoxide (\chem{CO}) or volatile organic substances (VOCs) and hydrocarbons (e.g., \chem{CH_4}) in the presents of nitrogen oxides (\chem{NO_x}). A typical reaction cylce for \chem{CO} is sketched below. In a sequence of rapid reactions a peroxy radical \chem{HO_2^\bullet} is formed through an initial reaction of \chem{CO} with a hydroxyl radical \chem{^\bullet OH}

@@

Via a reaction between \chem{HO_2^\bullet} and \chem{NO}, \chem{NO_2} is formed which is then photolysed. The resulting atomic oxygen reacts then with \chem{O_2} to form an ozone molecule

@@

  ^\bullet\chem{O(^3P)} + \chem{O_2} \rightarrow \chem{O_3}.

@@

A part from this, additional reaction pathways involving volatile organic compounds (VOCs) exist. Another possible source of tropospheric ozone is downward transport from the stratosphere via stratosphere-troposphere exchange (STE) \citep{WMO2014}. Major sinks are photolysis to \chem{O({^1{D}})} followed by a reaction with water vapor, reaction with \chem{HO_2} \citep{ACP:Seinfeld2006}, and de-nitration via

@@

  \chem{NO} + \chem{O_3} \rightarrow \chem{NO_2} + \chem{O_2}

@@

and

@@

  \chem{NO_2} + \chem{O_3} \rightarrow \chem{NO_3} + \chem{O_2},

@@

and scavenging processes, in particular dry deposition. A very special sink which is linked to dry deposition are so called ozone depleting events. They occure in spring-time in the polar boundary layer where an outburst of bromine monoxid \chem{BrO} (so called bromine explosion) leads to a rapid depletion of surface ozone \citep{JGR:Oltmans1981,GRL:Bottenheim1986,Nat:Barrie1988,JGR:Bottenheim2006}. Various schemes have been sucessfully applied to different types of atmospheric models \citep[e.g.][]{ACP:Yang2010,ACP:Toyota2011,ACP:Toyota2014a,GMD:Falk2018}. Although these events are important to understand surface ozone abundance in Arctic regions, we have not implemented any parameterization of these processes, yet.\\
Since ozone is highly reactive, its global mean life-time in the troposphere is roughly 22 days but ranges between a few days in the tropical boundary layer to up to one year in the upper troposphere \citep{JGR:Stevenson2005,ACP:Young2013}. The abundance of tropospheric ozone therefore varies, e.g., with time of the day (maximum $\sim$15:00 local time), season (mid-June maximum), altitude, and location \citep{ACP:Schnell2015} or weather conditions in general \citep{ACP:Otero2018}. Typical concentrations of surface ozone range from 10\,\unit{ppb} over the tropical Pacific to 100\,\unit{ppb} in the downwind areas of highly emitting sources \citep[Chapter 8]{IPCC2013}. This variability poses a challenge on both, trend analysis from observation as well as validation and intercomparison of models. At the observational side, there is only a limited number of long-term ozone observations, mainly restricted to European sites. Amongst these, only three display a statistical significant trend, indicating a doubling of tropospheric ozone since the onset of the industrialization \citep[Chapter 2]{IPCC2013}. But especially the very low pre-industrial ozone abundance cannot be reproduced by the likes of most models, while there is a general tendency to underestimate tropospheric ozone among the participating models in the Atmospheric Chemistry and Climate Model Intercomparison Project (ACCMIP) \citep[Chapter 8]{IPCC2013}.\\
Removal of any gas through gravitational settling or by uptake by plants and soil is referred to as dry deposition. The process of dry deposition is important for predicting surface ozone concentrations and understanding the observed amount and increase of tropospheric background ozone. Dry deposition velocities $v^i_d$ (also refered to as contuctances $G^i$) for any gaseous specie $i$ are typically calculated following a resistance-analogous approach

@@

wherein aerodynamic $R_a$, quasi laminar $R^i_b$, and canopy resistances $R^i_c$ are key components \citep{AE:Wesely1989}. For all gases $R_a$ is the same, while $R^i_b$ and $R^i_c$ vary from gaseous specie to specie and also depend on landuse types (e.g., ice/snow, water, urban, desert, agricultural land, deciduous forest, coniferous forest etc.). In its original formulation, fixed seasonal average dry deposition resistances have been used for each landuse type. For all three types of resistances in this Wesely-type parameterization, more process-oriented formulations have been developed and validated over the years, e.g., dry deposition of ozone to ocean \citep{ACP:Luhar2017} and stomatal conductance as part of canopy conductance \citep{PP:Buckley2017} or updates on the parameters have been provided from model validation with observation \citep[e.g., $v^\chem{O_3}_{ice/snow} = 1/10000\,\unit{cm s^{-1}}$,][]{ACP:Helmig2007}.

What is the status and what are the challenges in the dry deposition of ozone.

Impact of ocean? Although it is a low number there is 2\/3 ocean on earth... (Think, I read this in Hardacre paper... -> Luhar)

\section{Model description}
General model description followed by detailed dry deposition scheme \citep{ACP:Simpson2012,ICP:MappingManual2017}. This is followed by the necessary pre-processing of files. Old scheme \citep{AE:Wesely1989, JGR:Hough1991}

@@

\citet{GMD:Sovde2012,GMD:Lund2018}
\subsection{Ozone dry deposition scheme}
\label{subsec:DryDep}
Convention on Long-Range Transboundary Air Pollution (CLRTAP) -> European Monitoring and Evaluation Programme (EMEP),
simple plant function type (PFT)-based scheme. In this scheme, an average canopy resistance is computed from stomatal conductance of fractional PFTs in a grid cell.
\citep{ACP:Simpson2012} \citet{ICP:MappingManual2017}
\subsection{Greening season and photoactive radiation}

\section{Evaluation}

\subsection{Sensitivity tests}
Include table of all sensitivity runs.
\citep[e.g., $v^\chem{O_3}_{ice/snow} = 1/10000\,\unit{cm s^{-1}}$,][]{ACP:Helmig2007}
dry dep velocity over water \citep{JGR:Helmig2012}

\subsection{Comparison with recent multi-model evaluation}
Compare to \citet{ACP:Hardacre2015} as well as \citet{ACP:Luhar2017}.

\subsection{Comparison with observations}
The sites used in \citet{ACP:Hardacre2015} and additional ozone zonal data from satellites and sondes (\url{http://bodekerscientific.com/}).

>>>>

\begin{abstract}
  Since the industrial revolution, tropospheric background ozone concentrations have been increasing strongly mainly in the northern hemisphere. In recent years, the number of episodes of peak concentrations especially in summertime has been decreasing in North America and Europe due to the implementation of air quality regulations. At the same time, fast developing countries, like e.g., China or India, saw a significant increase in ozone related air pollution. High concentrations of ozone in ambient air are hazardous not only to humans but to the ecosystem in general. The impact of ozone damage on vegetation and agricultural plants in particular may even affect food security in the future. Most chemistry-transport models (CTMs) and chemistry-climate models (CCMs) tend to overestimate surface ozone values. This may be partly due to uncertainties in the distribution and amount of precursor species and stratosphere-troposphere exchange as well as scavenging processes removing ozone from the atmosphere. Removal of any gas through gravitational settling or by uptake by plants and soil is referred to as dry deposition. The process of dry deposition is important for predicting surface ozone concentrations and understanding the observed amount and increase of tropospheric background ozone. Dry deposition velocities are typically calculated following a resistance-analogous approach wherein aerodynamic, quasi laminar, and canopy resistances are key components, but these conceptual dry deposition velocities are hard to measure explicitly. Following the European Monitoring and Evaluation Programme (EMEP) MSC-W chemistry transport model, we compute canopy resistance from stomatal conductance. In this paper, we present an update on the dry deposition scheme implemented in the OsloCTM3v1.0 and evaluate the model with respect to ozone and different parameters in the stomatal conductance parameterization, e.g., photosynthetic photon flux density (PPFD) and phenology. Furthermore, we investigate the impact on modeled ozone concentrations, both at the surface and in the free troposphere. We show that ozone dry deposition in the new version (v.1.1) decreases with respect to the previous version (v1.0), leading to an increase in surface ozone of up to ??\,\unit{\%}. We also put our results in context to recent multi-model studies as well as satellite data. {\bf TODO: Finsish the results section as soon as there are results.}
\end{abstract}

@@

Tropospheric ozone is mainly produced in situ in complex photochemical cycles involving precursor gases such as carbon monoxide (\chem{CO}) or volatile organic substances (VOCs - also known as hydrocarbons) in the presents of nitrogen oxides (\chem{NO_x}). A typical reaction mechanism for \chem{CO} is sketched below. In a sequence of rapid reactions a peroxyl radical \chem{HO_2^\bullet} is formed through an initial reaction of \chem{CO} with a hydroxyl radical \chem{^\bullet OH}

@@

Via a reaction between \chem{HO_2^\bullet} and \chem{NO}, \chem{NO_2} is formed which is then photolysed. The resulting atomic oxygen reacts then with \chem{O_2} (and any other available co-reactant $M$) to form an ozone molecule

@@

  ^\bullet\chem{O(^3P)} + \chem{O_2} + M \rightarrow \chem{O_3} + M.

@@

Similar reaction mechanisms involving volatile organic compounds (VOCs) exist \citep{ACP:Monks2015}. Another source of tropospheric ozone is downward transport from the stratosphere via stratosphere-troposphere exchange (STE) \citep{WMO2014}. Based on observations, STE might only amount to roughly 10\,\unit{\%} ($550 \pm 140$\,\unit{Tg a^{-1}}) of the total global ozone budget in the troposphere, while ozone from chemical production is estimated to be 5000\,\unit{Tg a^{-1}} \citep{ACP:Monks2015}. Ozone is removed from the atmosphere by photochemical reactions or scavenging processes. Major sinks are photolysis followed by a reaction with water vapor to from \chem{OH}

@@

  \chem{O_3} + h\nu \rightarrow \chem{O(^1D)} + \chem{O_2})\\
  \chem{O(^1D)} + \chem{H_2O} \rightarrow 2\chem{OH},

@@

reactions with \chem{HO_2} \citep{ACP:Seinfeld2006}

@@

  {\bf TODO: Reaction},

@@

or de-nitration reaction via
  \chem{NO_x} + \chem{O_3} \rightarrow \chem{NO_{x+1}} + \chem{O_2}.
A minor sink which is related to dry deposition are so called ozone depleting events. They occur in spring-time in the polar boundary layer where an outburst of bromine monoxide \chem{BrO} (so called bromine explosion) leads to a rapid depletion of surface ozone \citep{JGR:Oltmans1981,GRL:Bottenheim1986,Nat:Barrie1988,JGR:Bottenheim2006}. Various schemes ranging from bulk-snow parameterizations \citep{ACP:Toyota2011,GMD:Falk2018} to detailed in-snow \citep{ACP:Toyota2014a}, and aerosol chemistry \citep{ACP:Yang2010} have been successfully applied to different types of atmospheric models. Although these events are important to understand surface ozone abundance in Arctic regions, we have not implemented any parameterization of these processes in the OsloCTM3.\\
Since ozone is highly reactive, its global mean life-time in the troposphere is roughly 22 days but ranges between a few days in the tropical boundary layer to up to one year in the upper troposphere \citep{JGR:Stevenson2005,ACP:Young2013}. The abundance of tropospheric ozone therefore varies, e.g., with time of the day (maximum $\sim$15:00 local time), season (mid-June maximum), altitude, location \citep{ACP:Schnell2015}, or weather conditions in general \citep{ACP:Otero2018}. Typical concentrations of surface ozone range from 10\,\unit{ppb} over the tropical Pacific to 100\,\unit{ppb} in the downwind areas of highly emitting sources \citep[Chapter 8]{IPCC2013}. This variability poses a challenge on both, trend analysis from observation as well as validation and intercomparison of models. At the observational side, there is only a limited number of long-term ozone observations, mainly restricted to European sites. Amongst these, only three display a statistical significant trend, indicating a doubling of tropospheric ozone since the onset of the industrialization \citep[Chapter 2]{IPCC2013}. But especially the very low pre-industrial ozone abundance cannot be reproduced by the likes of most models, while there is a general tendency to underestimate tropospheric ozone among the participating models in the Atmospheric Chemistry and Climate Model Intercomparison Project (ACCMIP) \citep[Chapter 8]{IPCC2013}.\\
Removal of any gas through gravitational settling or by uptake by plants and soil is referred to as dry deposition. The process of dry deposition is important for predicting surface ozone concentrations and understanding the observed amount and increase of tropospheric background ozone. It is estimated that about $1000 \pm 200$\,\unit{Tg a^{-1}} of ozone are removed from the atmosphere by dry deposition processes \citep{ACP:Monks2015}. Conceptually, dry deposition is a product between surface ozone concentration \chem{[O_3](h_0)} and a dry deposition velocity $v^\chem{O_3}_d$. Dry deposition velocities $v^i_d$ (also referred to as conductance $G^i$) for any gaseous specie $i$ are typically calculated following a resistance-analogous approach

@@

wherein aerodynamic $R_a$, quasi laminar $R^i_b$, and canopy resistances $R^i_c$ are key components \citep{AE:Wesely1989}. For all gases $R_a$ is the same, while $R^i_b$ and $R^i_c$ vary from gaseous specie to specie and also depend on landuse types (e.g., ice/snow, water, urban, desert, agricultural land, deciduous forest, coniferous forest etc.). Originally, \citet{AE:Wesely1989} used fixed seasonal average dry deposition resistances for each landuse type. For all three types of resistances in this Wesely-type parameterization, more process-oriented formulations have been developed and validated over the years, e.g., dry deposition of ozone to ocean \citep{ACP:Luhar2017} and stomatal conductance as part of canopy conductance \citep{PP:Buckley2017} or updates on the parameters have been provided from model validation with observation \citep[e.g., $v^\chem{O_3}_\text{ice/snow} = 1/10000\,\unit{cm s^{-1}}$,][]{ACP:Helmig2007}. The variaty of differing formulations leads to a wide spread in model intercomparisons \citet{ACP:Hardacre2015} and about 20\,\unit{\%} uncertainty on the resulting total dry deposition.\\

In Section~\ref{sec:model_des}, we will briefly describe the OsloCTM3, give a detailed account of the updates of the dry deposition scheme (Section~\ref{subsec:DryDep}) as well as present the pre-processing of meteorological input data to compute necessary input to the dry deposition scheme such as begin and duration of greening season and photosynthetic photon flux density (Section~\ref{subsec:pre-pro}). In Section~\ref{sec:eval}, we present sensitivity tests with respect to a manifold of parameters in the drydeposition scheme and validate our results with results from the modelintercomparison of \citet{ACP:Hardacre2015} and to observations.
{\bf TODO: Outlook on paper.}

\section{Model description}
\label{sec:model_des}
General model description followed by detailed dry deposition scheme \citep{ACP:Simpson2012,ICP:MappingManual2017}. This is followed by the necessary pre-processing of files. Old scheme \citep{AE:Wesely1989, JGR:Hough1991}

@@

\citet{GMD:Sovde2012,GMD:Lund2018}

\subsection{Ozone dry deposition scheme}
\label{subsec:DryDep}
Convention on Long-Range Transboundary Air Pollution (CLRTAP) -> European Monitoring and Evaluation Programme (EMEP),
simple plant function type (PFT)-based scheme. In this scheme, an average canopy resistance is computed from stomatal conductance of fractional PFTs in a grid cell.
\citep{ACP:Simpson2012} \citet{ICP:MappingManual2017}

\subsection{Greening season and photosynthetic photon flux density}
\label{subsec:pre-pro}

\section{Evaluation}
\label{sec:eval}

\subsection{Sensitivity tests}
\label{subsec:sens}
Include table of all sensitivity runs.
\citep[e.g., $v^\chem{O_3}_{ice/snow} = 1/10000\,\unit{cm s^{-1}}$,][]{ACP:Helmig2007}
dry dep velocity over water \citep{JGR:Helmig2012} Impact of ocean? Although it is a low number there is 2\/3 ocean on earth... (Think, I read this in Hardacre paper... -> Luhar)

\subsection{Comparison with recent multi-model evaluation}
\label{subsec:model}
Compare to \citet{ACP:Hardacre2015} as well as \citet{ACP:Luhar2017}.

\subsection{Comparison with observations}
\label{subsec:obs}
The sites used in \citet{ACP:Hardacre2015} and additional ozone zonal data from satellites and sondes (\url{http://bodekerscientific.com/}).
